Weighted Model Counting with Conditional Weights for Bayesian Networks

### 2021-06-02 doc/long_talk/talk.tex
Slides are almost ready

\documentclass{beamer}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{tikz}
\usepackage{minibox}
\usepackage{listings}
\usepackage{complexity}

@@

% BAs, pseudo-Boolean functions, measures, equivalence of definitions
\begin{frame}{An Alternative Way to Think About It}
\end{frame}

% The theorems: WMC is not expressive enough to cover all measures but can be
% extended to do so
\begin{frame}{Some Theoretical Results}
\end{frame}

% Things I could mention:
% The details of the transformation
% Experimental results

\begin{frame}{The Big Picture}
  \begin{columns}
    \begin{column}{0.35\textwidth}
      \centering
      \structure{Probabilistic Inference} \\
      (e.g., a Bayesian network)
      \fbox{
        \begin{minipage}{2.5cm}
          \centering
          \begin{tikzpicture}[edge from parent/.style={draw,-{Latex}}]
            \node[draw,circle] at (0, 0) (a) {$A$}
            child {node[draw,circle] (b) {$B$}};
          \end{tikzpicture}\\
          Find \alert{$\Pr(B = 0)$}
        \end{minipage}
      }
    \end{column}
    \begin{column}{0.35\textwidth}
      \centering
      \structure{WMC Encoding} \\
      (CNF with literal weights)
      \framebox{
        \scalebox{0.2}{
          \begin{tabular}{@{}l@{}}
            p cnf 8 17 \\
            -2 -1 0 \\
            1 2 0 \\
            -3 1 0 \\
            -1 3 0 \\
            -5 -1 0 \\
            -5 -4 0 \\
            1 4 5 0 \\
            -6 -1 0 \\
            -6 4 0 \\
            -4 1 6 0 \\
            -7 1 0 \\
            -7 -4 0 \\
            -1 4 7 0 \\
            -8 1 0 \\
            -8 4 0 \\
            -4 -1 8 0 \\
            -4 0 \\
            c weights 1.0 1.0 0.5 1.0 0.5 1.0 1.0 1.0 0.6 1.0 0.4 1.0 0.1 1.0 0.9 1.0
          \end{tabular}%
        }
      }

      \vspace{1cm}
      \onslide<2>{
      \structure{Conditional Weights} \\
      \framebox{
        \scalebox{0.2}{%
          \begin{tabular}{@{}l@{}}
            p cnf 2 1 \\
            2 0 \\
            w 1 0.5 0.5 \\
            w 2 1 0.6 0.4 \\
            w 2 -1 0.1 0.9
          \end{tabular}%
        }
      }
      }
    \end{column}
    \begin{column}{0.30\textwidth}
      \centering
      \structure{WMC Algorithm}\\
      \onslide<2>{\alert{$>100\times$} faster}
    \end{column}
  \end{columns}
  \begin{tikzpicture}[remember picture,overlay]
    \draw[-{Latex}] (3.4, 3) -- (4.9, 4);
    \draw<2>[-{Latex}] (3.4, 3) to [bend right] (5.8, 0.2);
    \draw[-{Latex}] (7.7, 4) -- (9, 2.8);
    \draw<2>[-{Latex}] (6.7, 0.2) to [bend right = 45] (9, 2.8);
  \end{tikzpicture}
\end{frame}
\end{document}

>>>>

\documentclass{beamer}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{tikz}
\usepackage{minibox}
\usepackage{listings}
\usepackage{complexity}
\usepackage{booktabs}

@@

\begin{frame}{An Alternative Way to Think About WMC}
  \begin{itemize}
  \item Let \structure{$V$} be the set of variables.
  \item Then \structure{$2^{2^V}$} is the Boolean algebra of propositional
    formulas.
  \end{itemize}
  \begin{definition}
    A \alert{measure} is a function \structure{$\mu\colon 2^{2^V} \to
      \mathbb{R}_{\ge 0}$} such that:
    \begin{itemize}
    \item \structure{$\mu(\bot) = 0$};
    \item \structure{$\mu(x \lor y) = \mu(x) + \mu(y)$} whenever \structure{$x
        \land y = \bot$}.
    \end{itemize}
  \end{definition}
  \begin{block}{Observation}
    WMC corresponds to the process of calculating the value of
    \structure{$\mu(x)$} for some \structure{$x \in 2^{2^V}$}.
  \end{block}
\end{frame}

% \begin{frame}{How  to Define a Measure}
%   \begin{itemize}
%   \item A \alert{weight function} is any function \structure{$\nu\colon 2^V \to
%       \mathbb{R}_{\ge 0}$}.
%     \begin{itemize}
%     \item It is \alert{factored} if
%       \[
%         \nu = \prod_{x \in V} \nu_x
%       \]
%       for some functions \structure{$\nu_x\colon 2^{\{x\}} \to \mathbb{R}_{\ge
%           0}$}, \structure{$x \in V$}.
%     \end{itemize}
%   \item We say that \structure{$\nu$} \alert{induces} \structure{$\mu$} if
%     \[
%       \mu(x) = \sum_{\{u\} \le x} \nu(u)
%     \]
%     for all \structure{$x \in 2^{2^V}$}.
%   \item A measure \structure{$\mu$} is \alert{factorable} if there exists a
%     factored weight function \structure{$\nu$} that induces \structure{$\mu$}.
%   \end{itemize}
% \end{frame}

\begin{frame}{The Limitations and Capabilities of Classical WMC}
  \begin{alertblock}{Observation}
    Classical WMC is only able to evaluate \structure{factorable} measures
    (c.f., a collection of mutually independent random variables).
  \end{alertblock}
  \begin{theorem}[Informal Version]
    It is always possible to add more variables to turn a non-factorable measure
    into a factorable measure.
  \end{theorem}
  However, that is not necessarily a good idea!
\end{frame}

\begin{frame}{Conditional Probability Tables (CPTs) as Functions}
  \begin{itemize}
  \item \alert{One} variable for every random variable with \alert{two} values.
  \item \alert{$k$} variables for every random variable with \alert{$k > 2$}
    values.
  \item Define indicator functions of the form \structure{$[x]\colon 2^{\{x\}}
      \to \{ 0, 1 \}$}.
    \begin{itemize}
    \item \structure{$[x](\emptyset) = 0$};
    \item \structure{$[x](\{ x \}) = 1$}.
    \end{itemize}
  \item Define \structure{$+$}, \structure{$\cdot$}, and scalar multiplication pointwise.
  \item Then a CPT can be represented as a function.
  \end{itemize}
  \begin{columns}[t]
    \begin{column}{0.5\textwidth}
      \centering
      \begin{tabular}[t]{ccc}
        \toprule
        $a$ & $b$ & $\Pr(A = a \mid B = b)$ \\
        \midrule
        1 & 1 & 0.6 \\
        1 & 0 & 0.4 \\
        0 & 1 & 0.1 \\
        0 & 0 & 0.9 \\
        \bottomrule
      \end{tabular}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{align*}
        \mathsf{CPT_A} &= 0.6[\lambda_{A=1}] \cdot [\lambda_{B=1}] \\
                       &+ 0.4[\lambda_{A=1}] \cdot \overline{[\lambda_{B=1}]} \\
                       &+ 0.1\overline{[\lambda_{A=1}]} \cdot [\lambda_{B=1}] \\
                       &+ 0.9\overline{[\lambda_{A=1}]} \cdot \overline{[\lambda_{B=1}]},
      \end{align*}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Experimental Results}
  \centering
  \input{cumulative}
\end{frame}

% \begin{frame}{Compared Against the Second-Best Encoding}
%   \centering
%   \input{scatter2}
% \end{frame}

\begin{frame}{Comparison With the State of the Art}
  \centering
  \input{scatter1}
\end{frame}

\begin{frame}{Summary and Future Work}
  \begin{itemize}
  \item Classical WMC can represent any probability distribution by adding more
    variables.
  \item But this is not the right approach for WMC algorithms that support
    working directly with functions.
  \item Specifically with ADDMC, avoiding redundant variables resulted in
    \alert{$127$} times faster inference.
  \item Could this idea be successfully applied to other applications of WMC or,
    perhaps, other WMC algorithms?
  \end{itemize}
  \begin{itemize}
  \item Potential improvements to the encoding:
    \begin{itemize}
    \item Apply ideas from other WMC encodings for Bayesian networks (e.g.,
      prime implicants, $\log$ encoding).
    \item Develop encoding tricks that apply to functions but not to conjunctive
      normal form.
    \item More on this in our SAT~2021 paper \emph{\structure{Weighted Model
          Counting Without Parameter Variables}}.
    \end{itemize}
  \end{itemize}
\end{frame}

\end{document}
